---
title: "Amsterdam House Price Analysis and Prediction"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
always_allow_html: true
fontsize: 16pt
urlcolor: blue
editor_options:
  markdown:
    wrap: 72
---

<style>
  body {
    margin: 4cm; /* Adjust the margin as needed */
  }
</style>

<hr style="border: 0; height: 1px; background-color: black;">
<hr style="border: 0; height: 3px; background-color: black;">


<center>
# <strong> From Data to Dwellings: <br>
# Decoding Amsterdam's Housing Prices </strong>


<span style="font-size: 15pt;">
 <b>Ayushman Anupam </b> <br>
<b>Chennai Mathematical Institute</b>
</span>
</center>

<hr style="border: 0; height: 3px; background-color: black;">

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE)
# Suppress package startup messages
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggplot2))
```
<br>

## <strong> Introduction </strong>
<hr style="border: 0; height: 2px; background-color: grey;">


This project aims to analyze and predict **housing prices in Amsterdam**
using comprehensive data collected in August 2021. The Amsterdam housing
market has experienced significant fluctuations in recent years, driven
by various factors such as economic conditions, demographics, and
housing policies. Understanding the dynamics of this market is crucial
for buyers, sellers, and investors alike. Therefore, the primary
objective of this analysis is to **identify and comprehend the trends
that influence housing prices in Amsterdam.** By examining a rich
dataset that includes detailed information about house prices and their
associated features, we seek to uncover variables that exhibit a strong
correlation with housing prices.

In our exploration of the dataset, we will focus on identifying key
predictors of housing prices. These predictors may include various
attributes such as the area of the property, the number of rooms, the
location's longitude and latitude, and other relevant features. Through
exploratory data analysis (EDA), we will **visualize these relationships
to determine which factors most significantly impact housing prices.**

Furthermore, the project will assess whether the identified predictors
can be effectively employed in a predictive model to estimate housing
prices and forecast market trends. By leveraging statistical techniques
and machine learning algorithms, we aim to develop a robust model that
offers a good fit for predicting housing prices based on the available
dataset. This model will not only facilitate a deeper understanding of
how various factors interact and contribute to price fluctuations but
also provide practical applications for stakeholders in the real estate
market.



```{r loading required libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(plotly)
library(RColorBrewer)
library(reshape2)
```

```{r load libraries and clean, message=FALSE, warning=FALSE, include=FALSE}

## reading data
AmHouse <- read.csv("C://Users//ayush//Downloads//Amsterdam House Price Ananlysis and prediction//HousingPrices-Amsterdam-August-2021.csv")
# Count the number of rows with NA values
NA_rows_count <- sum(!complete.cases(AmHouse))

```
<br><br>


## <strong> Data set Description </strong>
<hr style="border: 0; height: 2px; background-color: grey;">

This
[dataset](https://www.kaggle.com/datasets/thomasnibb/amsterdam-house-price-prediction/data?select=HousingPrices-Amsterdam-August-2021.csv)
detailed information about house prices in Amsterdam, Netherlands as of
August 2021.

The housing prices have been obtained from Pararius.nl as a snapshot in
August 2021. The original data provided features such as price, floor
area and the number of rooms. The data has been further enhanced by
utilising the Mapbox API to obtain the coordinates of each listing.

The Amsterdam House Price Prediction data set **contains 924** records
and includes features such as Address, Zip, Rooms, Area, Lat, Lon and
Price as defined below. However, there are **4 missing values** in the
"Price" field. To ensure data integrity, these **records with missing
prices are removed** before further analysis or modeling. This step
ensures accurate predictions by eliminating incomplete data points,
which could skew the results of the machine learning models designed to
predict house prices based on the remaining features.

The 7 for sale in and around Amsterdam as in the data set are:

-   **Address :** Residential address

-   **Zip :** Residential Zip code.

-   **Price :** Residential price in Euros.

-   **Area :** Residential area of House in square meters.

-   **Room :** Number of rooms in house

-   **Lon :** Longitude coordinates of location of house

-   **Lat :** Latitude coordinates of location of house


<br>


### <strong> Our Amsterdam data set table looks like </strong>

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}


# Replace the Address column with dashes
AmHouse$Address <- ". .. ... ..... ....."

# Print the modified data frame
print(head(AmHouse))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Print the count of rows with NA values
# print(paste("Number of rows with NA values:", NA_rows_count))


# Print rows with NA values
# NA_rows <- AmHouse[!complete.cases(AmHouse), ]
# print(NA_rows)

# removing NA values and storing it in a new Data Frame
Amsterdam_House <- na.omit(AmHouse)

# Reset row names
row.names(Amsterdam_House) <- NULL


# Again counting the number of rows with NA values to confirm 
NA_rows_count_new <- sum(!complete.cases(Amsterdam_House))

# Print the count of rows with NA values
#print(paste("Number of rows with NA values:", NA_rows_count_new))

# NOtE we don't need address and zip for our Data Analysis
# Remove the 'Address' column
Amsterdam_House <- Amsterdam_House %>%
  select(-Address, -Zip)

# Change column names
colnames(Amsterdam_House) <- c("Sl No", "Price", "Area", "Rooms", "Longitude", "Latitude")

```


<br><br>

## <strong> Exploratory Data Analysis </strong>
<hr style="border: 0; height: 2px; background-color: grey;">


### <strong> Figure 01. Histogram of (a) Price, (b) number of rooms and (c) Area of House</strong>

```{r echo=FALSE, fig.height=5, fig.width=12}

plot1 <- ggplot(Amsterdam_House, aes_string(x = "Price")) +
  geom_histogram(aes(y = ..density..), binwidth = 200000, fill = "purple", color = "black", alpha = 0.7) +
  geom_density(color = "red", size = 1) +
  ggtitle("Probability Distribution of House Prices") +
  labs(x = "Price (Euros)", y = "Density") +
  theme_minimal()+
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),         plot.background = element_rect(fill = "white", color = "black", size = 2))

plot2 <- ggplot(Amsterdam_House, aes_string(x = "Rooms")) +
  geom_histogram(aes(y = ..density..),fill = "blue",color = "black",alpha= 0.7,bins= 20) +
  geom_density(color = "red", size = 1, adjust = 3) +
  ggtitle("Probability Distribution ") +
  labs(x = "Number of rooms", y = "Density") +
  xlim(0,15 ) + 
  theme_minimal()+
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),         plot.background = element_rect(fill = "white", color = "black", size = 2))

plot3 <- ggplot(Amsterdam_House, aes_string(x = "Area")) +
  geom_histogram(aes(y = ..density..), binwidth =20,fill ="darkgreen",color ="black", alpha = 0.7,bins = 50) +
  geom_density(color = "red", size = 1) +
  ggtitle("Probability Distribution of House Area") +
  labs(x = "Area (in Sq meter)", y = "Density") +
  xlim(-10, NA) + 
  theme_minimal()+
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),         plot.background = element_rect(fill = "white", color = "black", size = 2))

grid.arrange(plot1, plot2, plot3, ncol = 3)
```

<br>

### <strong> Figure 02. (a) Number of Rooms vs Price of House and (b) Area vs Price of House </strong>

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.height =5, fig.width=12}
plot1 = ggplot(Amsterdam_House, aes(x = Rooms, y = Price)) +
  geom_point(color = "blue", size = 3) +  
  geom_smooth(method = "lm", col = "red", se = FALSE) +
  labs(title = "Rooms vs. Price", x = "Number of Rooms", y = "Price (in Euros)") + 
  theme_minimal()+
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),         plot.background = element_rect(fill = "white", color = "black", size = 2))

plot2 = ggplot(Amsterdam_House, aes(x = Area, y = Price)) +
  geom_point(color = "blue", size = 3) +  
  geom_smooth(method = "lm", col = "red", se = FALSE) + 
  labs(title = "Area vs. Price", x = " Area(in sq m)", y = "Price (in Euros)" ) + 
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),
        plot.background = element_rect(fill = "white", color = "black", size = 2))

grid.arrange(plot1, plot2, ncol = 2)
```

<br>

### <strong> Figure 03. (a) Longitude vs Price , (b) Latitude vs Price </strong>

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.height=4, fig.width = 12}
plot1 = ggplot(Amsterdam_House, aes(x = Longitude, y = Price)) +
  geom_point(color = "blue", size = 3) +  
  geom_smooth(method = "lm", col = "red", se = FALSE) + 
  labs(title = "Price vs. Logitude", x = "Longitude (in degree)",y = "Price (in Euros)") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),
        plot.background = element_rect(fill = "white", color = "black", size = 2))

plot2 = ggplot(Amsterdam_House, aes(x = Latitude, y = Price)) +
  geom_point(color = "green", size = 3) +  
  geom_smooth(method = "lm", col = "red", se = FALSE) + 
  labs(title="Price vs. Latitude",x="Latitude (in degree)",y="Price (in Euros)") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),
        plot.background = element_rect(fill = "white", color = "black", size = 2))

grid.arrange(plot1, plot2, ncol = 2)
```

<br>

### <strong> Figure 04. Plots between Area vs Rooms </strong>

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width = 12}
plot1 = ggplot(Amsterdam_House, aes(x = Rooms, y = Area)) +
  geom_point(color = "blue", size = 3) +  
  geom_smooth(method = "lm", col = "red", se = FALSE) + 
  labs(title = 'Area vs. Rooms', x = "Number of Rooms", y = "Area ( in sq m)")+
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),
        plot.background = element_rect(fill = "white", color = "black", size = 2))

plot2 = ggplot(Amsterdam_House, aes(x = as.factor(Rooms), y = Area)) +
  geom_boxplot(fill = 'cyan', color = 'black') +
  labs(title = 'Area vs. Rooms', x = "Number of Rooms", y = "Area ( in sq m)") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),
        plot.background = element_rect(fill = "white", color = "black", size = 2))

grid.arrange(plot1,plot2, ncol = 2)
```

<br>

### <strong> Figure 05. Plot of House Locations(longitude and Latitude), Price and Area </strong>

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=16}
ggplot(Amsterdam_House, aes(x = Longitude, y = Latitude)) +
  geom_point(aes(size = Price, color = Area), alpha = 0.7) +
  scale_color_gradient(low = 'blue', high = 'red') +
  labs(title = 'Plot of House Locations', x = 'Longitude', y = 'Latitude') +
  theme_minimal()+
  theme(
    panel.background = element_rect(fill = "white", color = "grey", size = 2),  
    plot.background = element_rect(fill = "white", color = "black", size = 2))
```

<br>

### <strong> Figure 06. Plot of Rooms vs Area coloured by Price </strong>

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.height=9, fig.width=16}
ggplot(Amsterdam_House, aes(x = as.factor(Rooms), y = Area, color = Price)) +
  geom_point(size = 4) +
  scale_color_gradient(low = 'yellow', high = 'red') +
  labs(title = 'Area vs. Rooms by Price', x = 'Rooms', y = 'Area') +
  theme_minimal()+
  theme(
    panel.background = element_rect(fill = "white", color = "grey", size = 2),        plot.background = element_rect(fill = "white", color = "black", size = 2))
```

<br>

### <strong>Figure 07. Interactive 3D plot of Area vs Room vs Price</strong>
<div style="border: 2px solid grey; padding: 20px; border-radius: 10px; background-color: #f9f9f9;">
<center>
```{r}
library(plotly)
plot_ly(Amsterdam_House, x = ~Area, y = ~Rooms, z = ~Price, 
        color = ~Price, colors = colorRamp(c("red", "blue", "green"))) %>%
  add_markers(size = 3) %>%
  layout(title = '3D Scatter Plot of Area, Rooms, and Price',
         scene = list(xaxis = list(title = 'Area'),
                      yaxis = list(title = 'Price'),
                      zaxis = list(title = 'Rooms')))
```
</center>
</div>

<br>

### <strong>Figure 08. Interactive 3D plot of Latitude vs Longitude vs Price </strong>
<center>
<div style="border: 2px solid grey; padding: 20px; border-radius: 10px; background-color: #f9f9f9;">
```{r}
library(plotly)
plot_ly(Amsterdam_House, x = ~Latitude, y = ~Longitude, z = ~Price, 
        color = ~Price, colors = colorRamp(c("red", "blue", "green"))) %>%
  add_markers(size = 3) %>%
  layout(title = '3D Scatter Plot of Latitude, Longitude, and Price',
         scene = list(xaxis = list(title = 'Latitude'),
                      yaxis = list(title = 'Longitude'),
                      zaxis = list(title = 'Price')))
```
</div>
</center>

<br>

## <strong>Correlation between Variables</strong>
<hr style="border: 0; height: 2px; background-color: grey;">
<center>
```{r Finding Correlatipn Matrix}

# Compute the correlation matrix for numeric variables
correlation_matrix <- cor(Amsterdam_House[, c("Price", "Area", "Rooms", "Longitude", "Latitude")])
```
</center>

<br>

### <strong> Figure 09. Heat map of correlation between fields of Amsterdam House data set </strong>
<center>
```{r plotting Correloation Heatmap, echo= FALSE, message=FALSE, warning=FALSE}
# Reshape the correlation matrix into long format
cor_melted <- melt(correlation_matrix)

# Convert Var1 and Var2 to character
cor_melted$Var1 <- as.character(cor_melted$Var1)
cor_melted$Var2 <- as.character(cor_melted$Var2)

# Filter to keep only the upper triangle, including the diagonal
cor_melted <- cor_melted[cor_melted$Var2 >= cor_melted$Var1, ]

# Define colors using RdYlBu palette
colors <- brewer.pal(n = 10, name = "RdYlBu")

# Create the heatmap with the specified aesthetics
heatmap_plot <- ggplot(cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(colors = colors, 
                       values = scales::rescale(c(-1, 0, 1)),  # Rescale for color mapping
                       na.value = "grey") +  # Color for NA values
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  theme_minimal() +
  labs(title = "Correlation Heatmap", x = "Fields", y = "Fields") +
  theme(axis.text.x = element_text( size = 15),  # Adjust text size
        axis.text.y = element_text(size = 15), 
        plot.title = element_text(hjust = 0.5, size = 15),  # Center the title
        axis.title.x = element_blank(),  # Remove x-axis title
        axis.title.y = element_blank()) +  # Remove y-axis title
  scale_x_discrete(position = "top") + # Move x-axis labels to the top
  theme(panel.background = element_rect(fill = "white", color = "grey", size = 2),
    plot.background = element_rect(fill = "white", color = "black", size = 2)) 
# Display the plot
print(heatmap_plot)

```

<br>

###  <strong>Correlation matrix for fields is </strong>

```{r}
print(correlation_matrix)
```
</center>

<br><br>

## <strong>Result </strong>
<hr style="border: 0; height: 2px; background-color: grey;">

The exploratory data analysis (EDA) conducted on the Amsterdam housing dataset provides valuable insights into the factors influencing housing prices. The key findings from the EDA are summarized below:

1.  **Univariate Analysis:**
\vspace{-2mm}
    -   **Price**: Histogram reveals a right-skewed distributed, with
        mean price around € 580,000.
    -   **Area**: Most properties range between 50 and 150 square
        meters.
    -   **Rooms**: Houses with 3 or 4 rooms are most common.
2.  **Bivariate Analysis:**
\vspace{-2mm}
    -   **Rooms vs. Price**: Strong positive correlation i.e, more rooms
        means high house price.
    -   **Area vs. Price**: positive correlation is observed i.e, larger
        homes are more expensive.
    -   **Price vs. Longitude/Latitude**: Weak correlation, but location
        still influences price.
    -   **Rooms vs. Area**: High correlation between room count and
        area.
3.  **Multivariate Analysis:**
\vspace{-2mm}
    -   **Price, Area, and Location**: Geographic location influences
        prices, though area remains a stronger predictor.
    -   **Rooms vs. Area by Price**: Higher-priced houses typically have
        more rooms and larger areas.
    -   **Correlation Heatmap**: Price is most strongly correlated with
        area and rooms, while longitude and latitude have weaker
        effects.

<br><br>

## <strong>Conclusion</strong>
<hr style="border: 0; height: 2px; background-color: grey;">
This project successfully analyzed and predicted housing prices in Amsterdam by examining a comprehensive dataset from August 2021. Through univariate, bivariate, and multivariate data analyses, we identified significant trends and correlations between various predictors such as area, number of rooms, and geographic location.

-   **Rooms and Area vs Price**: House price is strongly positvely
    influenced by area and the number of rooms, as seen in the various
    bivariate and multivariate plots. so, in prediction it will get more
    weight.

-   **Latitude and Longitude vs Price**: longitude and latitude, also
    affects house prices, though not as strongly as the size of the
    home. So, they will play a role in prediction but will have less
    weight.

-   **Correlation of Variables**: The correlation heatmap and pair plots
    reinforce the findings that area and room count are the strongest
    predictors of price, with location playing a secondary role.

    The strong positive correlation between the `Area` and `Price`, as
    well as the moderate correlation with `Rooms`, indicates that a
    linear relationship may exist between these variables and house
    prices. This suggests that a linear regression model could provide a
    solid baseline for predictions. Also, other predictors identified
    during the EDA , such as `Longitude` and `Latitude`, can be included
    as independent variables in the linear model. These features have
    shown significant relationships with house prices and are likely to
    contribute meaningfully to the model's accuracy.
    
    
    
    
<hr style="border: 0; height: 3px; background-color: black;">
<hr style="border: 0; height: 1px; background-color: black;">